---
title: "Exploring Mac Miller's Discography with Spotify API"
author: "Collin Smith"
date: '2022-05-19'
output: github_document
---

Hello, and welcome to my very first Markdown publication. I am still relatively new to R, with about 2 years of exposure and experience with the language. The vast majority of my time spent with R has been for university courses. Throughout this time I have learned a great deal from books like Hadley Wickham's [R for Data Science](https://r4ds.had.co.nz/), as well as tutorials, articles, and markdown pages from sites such as [RPubs](https://rpubs.com/), [R-bloggers](https://www.r-bloggers.com/), and [Towards Data Science](https://towardsdatascience.com/). My hope is that this project can in turn be used to help someone else down the road, as others have helped me.

In this project, we will be exploring how to use Spotify's API and the **spotifyr** package to access data about a particular artists discography. We'll also be utilizing the **genius** package to retrieve lyrics for songs, allowing for sentiment to be explored and visualized.

### Obtaining Spotify API Credentials

To begin, we will first need to head over to the [Spotify for Developers](https://developer.spotify.com/dashboard/) page, where we will be registering an application to obtain an API key. Once you've logged in, select "Create an App" and fill out the required fields. Completing this will give you access to two important fields: your **client id** and your **client secret** (or API key). These fields will be used to let the API know who is accessing it and that you have proper authentication. To get started, run the following code:

```{r auth-example, eval = F, include = T}
Sys.setenv(SPOTIFY_CLIENT_ID = 'your client id')
Sys.setenv(SPOTIFY_CLIENT_SECRET = 'your client secret')
access_token <- get_spotify_access_token()
```

```{r auth-actual, eval = T, echo = F}
source('auth_func.R')
```
